### 2022-03-18 paper.tex
a paper i wrote a while ago

What I propose is not, strictly, a unit of study, but rather, a focus that I would like to
see incorporated into the units already in your class. I find myself fascinated by the
economic realities of performance art, and the ways different systems of funding affect
the works produced. I was struck by this realization during a class from Brandon Woolf and
Andy Goldberg a few weeks ago, when we discussed Berlin's \textit{Freie Szene}. The norm
in Germany is for each theater to permanently employ an ensemble, who are paid a stable
salary by the state. The \textit{Freie Szene}, in contrast, consists of independent
troupes who are funded by government grants on a per-show basis. On the one hand,
\textit{freie} groups exist in a much more precarious space, where a series of rejected
proposals can leave them missing rent or collecting welfare. On the other hand, they enjoy
significant freedoms (hence the name) because of their disconnect from the institutional
hierarchies which employ more traditional ensembles. Notably, it is challenging for a
director working with a salaried ensemble to justify bringing in outside collaborators, so
temporarily hiring subject matter experts to collaborate on particular projects has become
a hallmark of the \textit{Freie Szene}. We got to see this play out onstage: \textit{Roma
  Armee}, a play interested in staging Romani performers, wound up with a cast that was
roughly half Romani, because the Gorki's ensemble only employed a handful of Romani
people. Rimini Protokoll's \textit{100\% Berlin}, on the other hand, invited and placed
onstage 100 Berlin residents. (In truth, I'm not sure if they were paid or volunteers, but
still.) I am interested in applying this same lens to the cultures of performance art we
discussed through the semester.

Throughout this essay, I will perform a relatively shallow investigation of various
economies of performance. I will compare these economies and their results to the
performance economy most familiar to me, Broadway (and American drama more broadly, as I
grew up in Minneapolis), but note that this is merely an artifact of convenience. My goal
is not to privilege Broadway as the optimal performance economy to which all history has
led; in fact, I think it's actually a pretty terrible version, and I find many of the
economies I'll discuss much more compelling.

Also note that most of what I write here is my own analysis based on my background
knowledge and what we discussed in class, along with a non-zero but frankly insufficient
amount of research. I caution you that what I write below will likely not reflect
scholarly consensus, because I'm not sure what the scholarly consensus is.\footnote{I
  blame this at least partially on the focus on novelty and reinterpretation in the
  humanities, which the computer scientist in me is unused to. In my discipline, it's not
  uncommon for the academic community to come to an agreement about the right way to do
  something, and then to stop worrying about it and move on to other things. This does not
  appear to be the case in theater studies or classics.}

\section*{5th century B.C.E. Athens}
% case study: greek drama

%% theater as a public service, performed by soldiers

Our class, like every theater history class, started with the Greeks, and so I suppose I
should, too. My understanding, based largely on a course I took with Peter Meineck to
satisfy my pre-1900s theater requirement, is that Athenian drama, was written by private
citizens for prestige, and performed by enlisted soldiers as part of their duties to the
city. I've also heard it proposed that wealthy Athenians would fund the playwrights, but
this was again with the aim of social prestige, not in exchange for artistic control or
seeking some return on investment. It's difficult to make broad statements about the
effects of this cultural organization on its products, as we have access to only a small
subset of the plays written by the three most successful playwrights, and none by any of
their myriad competitors. Still, we can trace some through-lines.

Athenian dramas tended to be quite specific to the year they were written and staged; I
think of \textit{Ajax} commenting on the rising tensions leading up to the Peloponnesian
war, and of \textit{Oedipus Tyrannus} responding to Pericles' death in the Plague of
Athens. Hyper-specific plays like this are discouraged in many other cultural economies,
including our own, by a desire that scripts be reusable and accessible. A Broadway show
which spoke deeply to its opening-night audience, but within a year started to alienate
all newcomers would be considered a failure, but the Athenian model allowed plays to do
just that.

Being performed as a public service, without a direct financial incentive like ticket
sales, seems to have encouraged Athenian playwrights to focus more on exposing their
audiences to new ideas or perspectives, and less on leaving them entertained or happy,
than the Broadway scene. I think once again of \textit{Ajax}, this time the ambiguous
light in which it portrays Ajax, one of Athens' founding heroes; Odysseus, arguably the
most beloved mythical hero; and Athena, the patron goddess of Athens. I'd love to tell you
what the Greek public thought of \textit{Ajax}, but all I can tell you is that there is no
record of its having won the City Dionysia. We do know that some very challenging and
provocative plays won first place, including Aeschylus' \textit{Persians}, which invited
sympathy for a hated enemy against whom Athens had quite recently fought and won a brutal
war. This suggests to me that the festival's convoluted judging scheme managed to select
for some definition of didacticism and artistic integrity, rather than pure entertainment
value.

\section*{Imperial Rome}

Roman theater, as I understand it, is best summarized by the phrase \enquote{bread and
  circuses.} Roman plays were funded either by the empire itself or by the patricians, and
served (with a range of conscious forethought) to placate and distract the plebeians and
to maintain the social hierarchy. At least as the post-Roman Christians tell it, this
meant the Romans' theater was gaudy, excessive and meaningless. Their main contribution to
drama as an art was in importing, preserving and re-performing the works of other
cultures; every surviving Greek play comes to us through the Romans. If I wanted to be
particularly political, I might set aside some time to draw parallels with modern America,
with the message that empire does not promote compelling art. That essay sounds like a
bummer, though, so instead I'll move on.

\section*{Medieval religious drama}

Frankly, I'm not that interested in this era, which is just as well because there's very
little to say about it. Medieval European history is characterized by the Church's
overwhelming influence over art, culture and day-to-day life. The Romans' excesses led the
Church to reject theater for the better part of a millennium, and when the art does
re-emerge, we see morality plays like \textit{Everyman}. It seems we actually know very
little about \textit{Everyman}; we don't know who wrote it, who first performed it, where
it was performed, or who paid for any of these things. I do feel safe to assert, though,
that the church was directly involved, either in a positive role by sponsoring the play's
creation or a negative one in restricting its content. Comparing this period with Broadway
theater is difficult, as our society is so secular compared to medieval Europe.

\section*{Shakespeare and the renaissance}

It's in the renaissance that we first see a theater economy similar to the one I live with
today, and by Shakespeare's time in the Elizabethan age, the parallels are
remarkable. Playwrights were commissioned by wealthy patrons, especially the monarchs, and
used that funding to support initial performances. They earned profit from ticket sales,
and used the prestige of popular shows to attract patrons for future works. The notable
difference compared to Broadway is that Shakespeare had no obligation to pay back his
patrons' gifts; they were not investors. But the model is fundamentally the same:
monarchs, aristocrats and other people of wealth used financial involvement to dictate the
content of new plays, and ticket sales provided an incentive to produce popular,
crowd-pleasing shows.

Patrons in this era used drama as a means of political conflict and propaganda. On the
world stage, this takes the form of the monarchs commissioning plays which glorify,
historicize and mythologize England's past. We can't actually say for certain which of
Shakespeare's plays were commissioned by Elizabeth or James, or at least, I can't find a
source which justifies the claim, but we do know that a disproportionate number of
Shakespeare's plays are devoted to England's frankly very boring history, and that the
monarchs were his primary patrons. More interesting, on a local level, we see at least one
instance of an internal dispute using a Shakespeare play to justify an attempted coup: the
Earl of Essex famously hired Shakespeare's company, the Chamberlain's Men, to perform
\textit{Richard II} in the hopes of swaying public opinion towards the would-be usurper
against Elizabeth. (It didn't work.)

The popular incentive of ticket sales mostly shows through in Shakespeare's dirty jokes
and reliance on convention. You've already read my take on the dirty jokes, though I'll
point out that they show up not only in \textit{Romeo and Juliet} but also in otherwise
serious tragedies; \textit{Theater Histories} cites the porter scene in
\textit{Macbeth}. In addition to saucy entertainment, audiences expected certain
conventions, especially in terms of plot and style. Comedies had to end with a marriage,
and tragedies with a death; \textit{Theater Histories} here points out that \textit{As You
  Like It} and \textit{Hamlet} had four each of the appropriate ending. Jeremy Lopez has a
whole book about the conventions expected by theater audiences at this time, and the way
playwrights (especially Shakespeare) sought to meet those expectations; he highlights puns
and wordplay, asides and soliloquy, and rapid tonal shifts to keep things exciting. Other
than the specific plot conventions, which have changed to accommodate a new zeitgeist, I
think this whole paragraph could apply to Broadway musicals (dramas tend to be more
experimental and less constrained); they're filled with crowd-pleasing plot moments like
the long-awaited kiss at the end, dirty jokes, songs which illustrate a character's
internal monologue, and pretty much every other device Lopez describes.

>>>>

\newcommand{\theplay}{\textit{Romeo and Juliet}}
\newcommand{\billy}{Shakespeare}
\newcommand{\theproduction}{\textit{Slapstick Shakespeare}}

I read \theplay{} in high school, led by an English teacher who frankly wasn't very
good. She taught the play as a serious tragic romance. I experienced it more as a dark
comedy; an elaborate farce at the expense of young people. I was aided by a good annotated
edition (I've unfortunately been unable to find the same one since), which explained all
the dirty jokes.

The plot of \theplay{} always struck me as being... less than romantic. The supposed
\enquote{star-crossed lovers} are a pair of very young teenagers (Juliet is 13) who become
instantly infatuated upon meeting at a party. In fact, it's not even the first time Romeo
has fallen hopelessly in love! I'd like to stage the play in such a way to highlight the
absurdity, bring the raunchy jokes to a forefront, and hopefully destroy the canonical
reading that it's a sincere romance.

The dramatic literature major in me feels obligated to describe my approach as a comedic
\textit{Verfremdungseffekt}. Similar to how Brechtian drama imposes distance in order to
force its audience to critically reconsider the social and political circumstances
presented, \theproduction{} will force its audience to critically reconsider the canonical
interpretation of \theplay{}, particularly in terms of genre and tone.

\section*{A new prologue}

Rather than a chorus, I think I'll read the two prologues (there's one before each of the
first two acts), wearing a mobcap with a feather, to signify my being in-character as the
Bard. Before the first prologue, I'll enter holding the hat, to signify being
out-of-character, and address the audience with an introduction to set the tone:

\begin{dialogue}
  \speak{Phoebe} Hey, y'all. I'm Phoebe, and I'd like to welcome you to this production of
  \theplay{}, \billy{}'s beautiful tragic romance. There will be one 15-minute
  intermission between the second and third acts. Please silence your phones if you
  haven't already, and unwrap your snacks now so they don't crinkle during the show.

  Before we start, I know the language of \billy{} can be hard to get used to, and people
  sometimes have a hard time following, so I'll give you a quick plot summary so you know
  what's going on.

  We start with some exposition, to show you that the houses of Montague and Capulet are
  enemies. It's boring; I'd skip that part if I could. Whatever.

  Romeo of Montague is a thirsty teen, depressed because his crush, Rosaline, doesn't like
  him back. His pals Mercutio and Benvolio sneak him into a party to cheer him up, where
  he meets 13-year-old Juliet of Capulet. They fall in love at first sight. \direct{rolls
    eyes and mimes making themself barf.}

  That night, Romeo sneaks around outside Juliet's house and spies on her a little. He
  hears her talk about how hot he is, so he does his best to act innocent and pretend he
  was just passing by, and reveals himself. They get married the next morning.

  Then, drama! Fiery Tybalt of Capulet kills Mercutio, Romeo kills Tybalt in revenge, and
  the prince comes along and exiles Romeo from the city.

  Juliet decides she doesn't really care that Romeo murdered her cousin, and arranges for
  him to visit that night for sex, before he leaves town forever.

  Then, more drama! Juliet's dad arranges a high-society marriage for her, not knowing
  that Juliet is already married and deflowered. Juliet arranges some convoluted escape
  plan which involves faking her death. Romeo thinks she's actually dead, and immediately
  kills himself out of grief. Juliet wakes up, discovers Romeo's corpse, and kills herself
  for real. Then there's some feel-good stuff where houses Montague and Capulet realize
  the cost of their feud and make peace, which if you ask me brings us dangerously close
  to saying \direct{with airquotes} \enquote{hey kid, kill yourself to solve your parents'
    problems!} But who asked me, right?

  \direct{checks watch.} Oh, shit, I've spent way too long already. Without further ado,
  \direct{dons hat and quickly reads the actual prologue}.
\end{dialogue}

I've also considered breaking this summary up into smaller chunks, which I'd read before
each scene or each act. This could have the added benefits of providing some entertainment
during set changes, and of further distancing the audience by interrupting their
immersion.

\section*{Modifications to the script}

If this staging is to convincingly attack the idea that \billy{} wrote \theplay{} as a
sincere romantic tragedy, it must present his script, unaltered, in its entirety. That
said, many of \billy{}'s best jokes rely on idioms which no longer make any sense. As a
result, a narrator will pause the play at strategic points to explain. (I imagine myself
serving as this narrator, and so I will write my own name in the script.) For example, in
Act 1, Scene 1, supplementing lines 18--24, I would add:

\begin{dialogue}
  \speak{Gregory} The quarrel is between our masters and us their men.

  \speak{Sampson} 'Tis all one, I will show myself a tyrant: when I \\
  have fought with the men, I will be cruel with the \\
  maids, and cut off their heads.

  \speak{Gregory} The heads of the maids?

  \speak{Sampson} Ay, the heads of the maids, or their maidenheads; \\
  take it in what sense thou wilt.

  \speak{Phoebe} \direct{while entering from offstage} Could you pause for a moment, guys?

  \direct{to audience} In \billy{}'s time, a \enquote{maiden} is specifically a virgin,
  and when Sampson talks about \enquote{cutting off their maidenheads,} he means fucking
  them.

  \direct{beat}

  \direct{to \refer{Sampson} and \refer{Gregory}} Thanks, guys, go ahead. \direct{exits}
\end{dialogue}

These interruptions would be kept to a minimum, in order to preserve the flow of the
dialogue. In most cases, strategic emphasis and body language should be sufficient to get
the point across. The lines directly following make for a good example:

\begin{dialogue}
  \speak{Gregory} They \direct{chuckles, then continues while wiggling his eyebrows at the
    audience} must take it in the sense that feel it.

  \speak{Sampson} \direct{mimes holding an exaggerated erect penis} \\
  Me they shall feel while I am able to stand: and \\
  'tis known I am a pretty piece of flesh.

  \speak{Gregory} 'Tis well thou art not if; if thou hadst, thou \\
  hadst been \direct{pauses to produce an ugly wooden fish and hold it to his crotch} poor-John.
\end{dialogue}

\section*{Setting}

One of the most powerful ways to impose distance between the stage and the audience is to
add inconsistencies, so that the play lacks a coherent internal world. This is convenient
because it is easy from a production standpoint, and often reads as comedic.

\theproduction{} will make no effort to ground itself in any one place or time. Visual
motifs in the set and costumes will be chosen with an eye for what best portrays the
correct tone or character. Examples to follow.

\section*{Set}

The set will be dirt-cheap, and will look ridiculous. As with all my choices, my goal here
is to prevent the audience from taking things too seriously.

The only set piece which I see as necessary is the obvious Juliet-style balcony. I haven't
fully decided just how vulgar I want \theproduction{} to be, but I can certainly imagine a
version where the Juliet balcony is shaped and painted to look like a vagina, to emphasize
the sexual subtext in Act 2 Scene 2 and Act 3 Scene 5. A set of pink drapes or blinds
could make this especially funny, with the drapes closed in Act 2 Scene 2 (before the
couple has consummated their marriage), and opened in Act 3 Scene 5 (after the fact). A
wide range of subtler presentations are available, depending on how vulgar I wanted to be;
anything from a simple white wall with pink finish and pink drapes, all the way to an
O'Keeffe painting. I suspect the actual decision would be made by what my investors and
theater administrators would let me get away with.

As to other sets and settings, I imagine:

\begin{itemize}
\item \enquote{A hall in Capulet's house,} where Romeo and friends crash a party and meet
  Juliet, will be styled like a 2000s or 2010s high school dance. There will be a disco
  ball, a punch bowl, and extras in ill-fitting rented suits dancing the \textit{Cha-Cha
    Slide}. The intention is to say: this is where teenagers go for romance and
  excitement.
\item \enquote{Friar Laurence's Cell} will be my best approximation of an authentic monk's
  cell, with a cot, a bench, a candelabra, an ornate cross and little else. The intention
  is to say: here is a man whose intense piety has no place alongside the gaudy wealth and
  teenaged angst of our protagonists.
\item \enquote{A churchyard; in it a tomb belonging to the Capulets} will be decorated
  like a cheap haunted house, with plastic gravestones and fake skeletons purchased at a
  \textit{Spirit Halloween} (or equivalent pop-up store). The intention is to say: this is
  not a serious encounter with morality, it is a morbid punchline.
\end{itemize}

\section*{Costumes}

In no particular order:

\begin{itemize}
\item Prince Escalus and Count Paris, who should appear regal, wealthy and pretentious,
  will wear medieval finery, with frilly shirts, capes, furs and the like. The prince will
  wear a crown and carry a scepter. The count will wear a smaller coronet.
\item Mercutio, who should appear friendly, brash and carefree, will wear jeans and a high
  school letterman jacket. He will carry an open can of beer at all times.
\item Lord Capulet, who should appear a stern businessman who knows that violence is bad
  for profits, will mimic Marlon Brando as the Godfather.
\item Lady Capulet, similarly, will mimic Morgana King as Mama Corleone.
\item Lord Montague will mimic Richard Conte as Emilio Barzini, since I think the
  \textit{Godfather} references are fun.
\item Fiery Tybalt, who's always looking for a fight, will be a greaser, with a leather
  jacket, a cigarette and a shiv.
\item Benvolio, who should appear kind of a wimp, will be a stereotypical nerd. I imagine
  Crispin Glover as George McFly in \textit{Back to the Future}, with thick glasses,
  slicked-back hair, and a dress shirt with the top button buttoned but no tie.
\item Friar Laurence will be a bald monk wearing a brown robe, for similar reasons as my
  design for his cell.
\item Romeo, who should appear angsty, young and stupid, will be a goth teen. Dyed
  jet-black hair, huge steel ear studs, a thin goatee, tight black ripped jeans and a tee
  shirt.
\item Juliet, who should appear young, stupid and naïve, will have long wavy brown hair, a
  pretty blue dress, and lots of blush.
\end{itemize}
